---
title: Covid 19 Charts
description: A first play with R, charts and UK government Covid-19 data.
output: html_document
category: covid
tags: [r, covid]
---


## Playing with UK Covid data

The source of all data in this script is the UK Government website, <https://coronavirus.data.gov.uk/>.

```{r echo = FALSE, include = FALSE}
library(tidyverse)
library(zoo)

theme_set(theme_bw())
```

```{r get data, include = FALSE}
cases <- read_csv("https://api.coronavirus.data.gov.uk/v2/data?areaType=ltla&metric=newDeaths28DaysByDeathDate&metric=newCasesBySpecimenDate&format=csv")
```

Now we can get a summary of the `cases` data:

```{r}
summary(cases)
```

This shows that we have *162382* observations, with a dates between `r min(cases$date)` and `r max(cases$date)`.

### Wrangling the Data

To aide simplicity, we first choose a selected number of areas for data to compare and wrangle the data slightly to provide what we''ll be needing.

```{r wrangle data}
selected <- cases %>%
  filter(areaName %in% c("Crawley", "Mid Sussex", "Horsham")) %>%
  select(areaName, date, newCasesBySpecimenDate, newDeaths28DaysByDeathDate) %>%
  mutate(
    cases7 = rollmean(newCasesBySpecimenDate, 7, na.pad = TRUE),
    deaths7 = rollmean(newDeaths28DaysByDeathDate, 7, na.pad = TRUE)
  )
```

1. Filter our areas.  Just for local interest in this overview, we will look at Crawley, Horsham and Mid Sussex.
1. Select the columns we are interested in.
1. Mutate the data to add columns for 7-day Rolling Average of cases and deaths.

#### Crudely Plot our Data

Using the 7 day rolling average, to avoid a jagged chart, we show deaths and cases.

```{r deaths by date, warning = FALSE}
selected %>%
  ggplot(aes(date, deaths7, colour = areaName)) +
  geom_line(na.rm = TRUE) +
  labs(y = "# Deaths (7 day rolling average)", x = "Death Date")
```

The `Death Date` is the date of death, given that they had **Covid-19** within the past 28 days.

```{r warning = FALSE}
selected %>%
  ggplot(aes(date, cases7, colour = areaName)) +
  geom_line()
```

### Chart for Cases and Deaths

Not the best way to display this data, but it shows the number of cases at a given time, and how the death rates followed the cases.

```{r 7 day cases and deaths, warning = FALSE}
selected %>%
  ggplot(aes(date, newCasesBySpecimenDate, colour = areaName)) +
  geom_line(aes(y = cases7)) +
  geom_line(aes(y = deaths7)) +
  scale_y_log10() +
  labs(
    y = "# Cases", x = "Case Specimin Date",
    title = "7 Day rolling average of cases and deaths in selected areas"
  )
```

A boxplot with jitter to show the 'range' of data.  Notice that each area have a similar mean and quartile data.  Also that even the outliers are similar.

```{r box plot cases, fig.cap = "Boxplot with jitter for cases in selected areas"}
selected %>%
  ggplot(aes(areaName, newCasesBySpecimenDate, colour = areaName)) +
  geom_boxplot(outlier.shape = NA) +
  labs(y = "# Cases", x = "Area", title = "Cases in selected areas") +
  geom_jitter(width = 0.3, alpha = 0.4, stroke = 0)
```

Maybe a better way would be to facet this data.  Unfortunately, at the moment, our data is not truely 'tidy'.  In order to facet wrap, we need to tidy the `cases7` and `deaths7` columns.

"Tidy" data follows these rules:

1. Each variable in the data set is placed in its own column
1. Each observation is placed in its own row
1. Each value is placed in its own cell

<https://garrettgman.github.io/tidying>

Oh.  We need to turn

areaName|cases|deaths
---|---|---
X|10|6

into

areaName|ra7|count_type
---|---|---
X|10|cases
X|6|deaths

so that each row provides us with a single observation.

We use `pivot_longer` to create columns `ra7` and `count_type` and populate it with death/case values and detail the data type.

```{r}
tidy <- selected %>%
  select(-newCasesBySpecimenDate, -newDeaths28DaysByDeathDate) %>%
  pivot_longer(c(cases7, deaths7), names_to = "count_type", values_to = "ra7")
```

```{r 7 day cases and deaths faceted, warning = FALSE}
tidy %>%
  ggplot(aes(date, ra7, colour = areaName)) +
  geom_line(na.rm = TRUE) +
  facet_wrap(~count_type) +
  geom_smooth(method = loess) +
  scale_y_continuous(trans = "log10") +
  labs(
    y = "# Cases", x = "Case Specimin Date",
    title = "7 Day rolling average of cases and deaths in selected areas"
  )
```

## Sum Across the UK

Just for fun, join everything together and sum cases across the UK...

```{r sum cases across UK, warning = FALSE}
cases %>%
  mutate(cases = rollmean(newCasesBySpecimenDate, 7, na.pad = TRUE)) %>%
  group_by(date) %>%
  summarise(sum_cases = sum(cases, na.rm = TRUE)) %>%
  ggplot(aes(date, sum_cases)) +
  geom_line() +
  scale_y_log10() +
  labs(
    y = "# Cases", x = "Case Specimin Date",
    title = "7 day RA of positive cases in the UK"
  )
```

Using a logarithmic scale, we can clearly see just how high cases are still - and how quickly it took hold in March / April 2020.

## Summary

There are so many (better and correct) ways to present this data.  These charts are just a tiny amount of fun for my first 'used in anger' R script.
