---
title: Covid-19 Charts
output: html_document
category: covid
tags: [r, covid]
---

# Playing with UK Covid data

```{r echo = FALSE, include = FALSE}
library(tidyverse)
library(zoo)
```

```{r get data, include = FALSE}
cases <- read_csv("https://api.coronavirus.data.gov.uk/v2/data?areaType=ltla&metric=newDeaths28DaysByDeathDate&metric=newCasesBySpecimenDate&format=csv")
```

Now we can get a summary of the `cases` data:

```{r}
summary(cases)
```

This shows that we have *162382* observations, with a dates between `r min(cases$date)` and `r max(cases$date)`.

## Wrangling the Data

To aide simplicity, we first choose a selected number of areas for data to compare and wrangle the data slightly to provide what we''ll be needing.

```{r wrangle data}
selected <- cases %>%
  filter(areaName %in% c("Crawley", "Mid Sussex", "Horsham")) %>%
  select(areaName, date, newCasesBySpecimenDate, newDeaths28DaysByDeathDate) %>%
  mutate(
    cases7 = rollmean(newCasesBySpecimenDate, 7, na.pad = TRUE),
    deaths7 = rollmean(newDeaths28DaysByDeathDate, 7, na.pad = TRUE)
  )
```

1. Filter our areas.
1. Select the columns we are interested in.
1. Mutate the data to add columns for 7-day Rolling Average of cases and deaths.

### Crudely Plot our Data

Using the 7 day rolling average, to avoid a jagged chart, we show deaths and cases.

```{r deaths by date, warning = FALSE}
selected %>%
  ggplot(aes(date, deaths7, colour = areaName)) +
  geom_line() +
  labs(y = "# Deaths (7 day rolling average)", x = "Death Date") +
  theme_classic()
```

The `Death Date` is the date of death, given that they had **Covid-19** within the past 28 days.

```{r}
selected %>%
  ggplot(aes(date, cases7, colour = areaName)) +
  geom_line() +
  theme_classic()

selected %>%
  ggplot(aes(date, cases7, fill = areaName)) +
  geom_bar(position = "fill", stat = "identity") +
  theme_classic()
```

## Chart for Cases and Deaths

Not the best way to display this data, but it shows the number of cases at a given time, and how the death rates followed the cases.

```{r 7 day cases and deaths, warning = FALSE}
selected %>%
  ggplot(aes(date, newCasesBySpecimenDate, colour = areaName)) +
  geom_line(aes(y = cases7)) +
  geom_line(aes(y = deaths7)) +
  #  geom_point() +
  #  geom_smooth(method = loess) +
  scale_y_continuous(trans = "log10") +
  labs(y = "# Cases", x = "Case Specimin Date", title = "7 Day rolling average of cases and deaths in selected areas at 07 May 2021") +
  theme_classic()
```

```{r box plot cases, fig.cap = "Boxplot with jitter for cases in selected areas"}
selected %>%
  ggplot(aes(areaName, newCasesBySpecimenDate, colour = areaName)) +
  geom_boxplot(outlier.shape = NA) +
  labs(y = "# Cases", x = "Area", title = "Cases in selected areas at 07 May 2021") +
  geom_jitter(width = 0.3, alpha = 0.4, stroke = 0) +
  theme_bw()
```
