---
title: Covid-19 Heatmaps
description: Heatmaps of Covid-19 cases in the UK; using data from public government data
output: html_document
category: covid
tags: [r, covid]
layout: post
---

# Let's Plot Covid-19 in the UK...

```{r load, include=FALSE}
rm(list = ls())

library(tidyverse)
library(readxl)
library(zoo)
library(cowplot)

# prevent scientific notation below 10000
options(scipen = 9999)

knitr::opts_chunk$set(
  cache = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.height = 40,
  fig.width = 30
)
```

## Source

### Covid-19

Covid data is from the UK Government website, using the following URL:

<https://api.coronavirus.data.gov.uk/v2/data?areaType=ltla&metric=newDeaths28DaysByDeathDate&metric=newCasesBySpecimenDate&format=csv>

Covid-19 data is CONFIRMED cases or deaths only.

- A case is registered on the date of the test specimen.
- A death is registered on the day of death, and the death occurs within 28 days of a positive Covid-19 test.
- The dates, especially more recent ones, are subject to change as data comes in.

### Population

This data is sourced from the [ONS](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/populationestimatesforukenglandandwalesscotlandandnorthernireland) website.  It is from mid 2019, but good enough for our use.

### Read and Select Required Data

```{r source, dependson="load"}
source <- "https://api.coronavirus.data.gov.uk/v2/data?areaType=ltla&metric=newDeaths28DaysByDeathDate&metric=newCasesBySpecimenDate&format=csv"
src <- read_csv(source)[, c(1:2, 4:6)]

source_counties <- "http://geoportal1-ons.opendata.arcgis.com/datasets/89109c7ba29f496187e125b5c8a091b6_0.csv"
counties <- read_csv(source_counties)[, c(1, 4)]

source_population <- "../data/population-mid2019.xls"
population <- read_excel(
  source_population,
  sheet = "MYE2 - Persons",
  skip = 4
)[, c(1, 4)]

raw <- src %>%
  left_join(counties, by = c("areaCode" = "LTLA18CD")) %>%
  left_join(population, by = c("areaCode" = "Code"))

colnames(raw) <- c(
  "code", "name", "date", "cases",
  "deaths", "county", "population"
)
```

## Tidy...

Create an R-tidy table for easier use within `ggplot`.

```{r tidy, dependson="source"}
data <- raw %>%
  pivot_longer(c("cases", "deaths"), names_to = "type", values_to = "count")
data
```

For simplicity, split the data into `cases` and `deaths`.

Add some helper columns to each dataset.

Convert case and death counts to a rolling 7-day average.

```{r cases-helpers, dependson="tidy"}
cases <- subset(data, type == "cases")
deaths <- subset(data, type == "deaths")

cases_heatmap <- cases %>%
  arrange(code, date) %>%
  na.omit() %>%
  group_by(code) %>%
  mutate(
    max_count = max(count, na.rm = TRUE),
    max_date = date[which(count == max_count)][1],
    count7 = rollmean(count, 7, na.pad = TRUE),
    prop_max = count / max_count,
    per_ht = as.integer(count7 * 100000 / population),
    total = sum(count, na.rm = TRUE),
    total_per_ht = as.integer(total * 100000 / population)
  ) %>%
  ungroup()

deaths_heatmap <- deaths %>%
  arrange(code, date) %>%
  na.omit() %>%
  group_by(code) %>%
  mutate(
    max_count = max(count, na.rm = TRUE),
    max_date = date[which(count == max_count)][1],
    count7 = rollmean(count, 7, na.pad = TRUE),
    prop_max = count / max_count,
    per_ht = as.integer(count7 * 100000 / population),
    total = sum(count, na.rm = TRUE),
    total_per_ht = as.integer(total * 100000 / population)
  ) %>%
  ungroup()

min_date <- min(cases$date)
max_date <- max(cases$date)
```

## Heatmap 7-day Rolling Average

If you zoom in with your browser, you should just about be able to read the axes.

Total cases for 7-day rolling average.

```{r cases-avg-heatmap, dependson="cases-helpers", echo=FALSE}
cases_heatmap %>%
  ggplot(aes(
    x = date,
    y = fct_reorder(name, max_date),
    fill = prop_max * 100
  )) +
  geom_tile(color = "white", show.legend = TRUE) +
  scale_fill_distiller(palette = "Spectral", name = "Avg Cases / Max (%)") +
  scale_x_date(
    limits = as.Date(c(min_date, max_date)),
    date_breaks = "2 months",
    date_labels = "%b-%y",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = "",
    expand = c(0, 0)
  ) +
  theme_classic() +
  labs(
    title = "Timeline for COVID-19 Cases in UK Local Authorities",
    subtitle = paste0("The heatmap represents the 7-day rolling average of the number of confirmed cases within each Local Authority (LA), values are normalised to the maximum value within each LA.
LAs are ordered by the date at which they recorded their maximum number of cases.
Data updated to ", max_date),
    caption = "Plot by @misstickles"
  ) +
  theme(
    axis.line.y = element_blank(),
    plot.title.position = "plot",
    legend.position = "bottom"
  )
```

Each LA has its number of cases normalised to its own maximum number of cases.  Therefore, each LA will have a red plot signifying the day of maximum cases.

This way, it is easy to spot the first, second, and the third waves (just beginning).

Some (very brief) interesting things:

- some of the LAs with later maximum case dates (at the top), are seeing very high rates in the 3rd wave (proportionally)
- the early starters in the first wave are also starting early in the third
- there are a number of LAs (those who saw the earliest cases) who had large waves in both October / November AND January

I like this plot, as it aditionally shows clear information on where the cases began in the UK.

## Introduce population

This similar heatmap presents the number of cases per 100,000 population within each Local Authority.  As before, LAs are ordered by the date a which they recorded their maximum number of cases.

On the right is a plot of the total number of cases recorded for that LA.

```{r cases-heatmap, dependson="cases-helpers", echo=FALSE}
case_tiles <- cases_heatmap %>%
  ggplot(aes(
    x = date,
    y = fct_reorder(name, max_date),
    fill = per_ht
  )) +
  geom_tile(color = "white", show.legend = TRUE) +
  scale_fill_distiller(palette = "Spectral", name = "Cases / day") +
  scale_x_date(
    limits = as.Date(c(min_date, max_date)),
    date_breaks = "2 months",
    date_labels = "%b-%y",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = "",
    expand = c(0, 0)
  ) +
  theme_classic() +
  labs(
    title = "Timeline for COVID-19 Cases in UK Local Authorities, per 100,000 population",
    subtitle = paste0("The heatmap represents the number of confirmed cases per 100,000 within each Local Authority (LA).
LAs are ordered by the date at which they recorded their maximum number of cases.
Data updated to ", max_date),
    caption = "Plot by @misstickles"
  ) +
  theme(
    axis.line.y = element_blank(),
    plot.title.position = "plot",
    legend.position = "bottom"
  )

case_cols <- cases_heatmap %>%
  filter(date == max_date) %>%
  ggplot(aes(
    x = total_per_ht,
    y = fct_reorder(name, desc(max_date)),
    fill = total_per_ht
  )) +
  geom_col(show.legend = TRUE) +
  scale_fill_distiller(palette = "Spectral", name = "Total") +
  scale_x_continuous(
    name = "Total cases",
    breaks = c(0, 5000, 10000, 15000),
    expand = c(0, 0)
  ) +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    legend.position = "bottom"
  )

plot_grid(case_tiles, case_cols, align = "h", rel_widths = c(1, 0.2))
```

The general pattern is the same as above.  It is less easy to see the first wave, but more easy to identify the regions with the higher number of cases.

## Deaths

As for cases above, the number of deaths is normalised to the maximum of deaths recorded for each LA.

```{r deaths-heatmap, dependson="cases-helpers", echo=FALSE}
death_tiles <- deaths_heatmap %>%
  ggplot(aes(
    x = date,
    y = fct_reorder(name, max_date),
    fill = prop_max * 100
  )) +
  geom_tile(color = "white", show.legend = TRUE) +
  scale_fill_distiller(palette = "Spectral", name = "Avg Deaths / Max (%)") +
  scale_x_date(
    limits = as.Date(c(min_date, max_date)),
    date_breaks = "2 months",
    date_labels = "%b-%y",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = "",
    expand = c(0, 0)
  ) +
  theme_classic() +
  labs(
    title = "Timeline for COVID-19 Deaths in UK Local Authorities",
    subtitle = paste0("The heatmap represents the 7-day rolling average of the number of deaths within each Local Authority (LA), values are normalised to the maximum value within each LA.
LAs are ordered by the date at which they recorded their maximum number of cases.
Data updated to ", max_date),
    caption = "Plot by @misstickles"
  ) +
  theme(
    axis.line.y = element_blank(),
    plot.title.position = "plot",
    legend.position = "bottom"
  )

death_cols <- deaths_heatmap %>%
  filter(date == max_date) %>%
  ggplot(aes(
    x = total_per_ht,
    y = fct_reorder(name, desc(max_date)),
    fill = total_per_ht
  )) +
  geom_col(show.legend = TRUE) +
  scale_fill_distiller(palette = "Spectral", name = "Total") +
  scale_x_continuous(
    name = "Total cases",
    breaks = c(0, 100, 200, 300, 400),
    expand = c(0, 0)
  ) +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    legend.position = "bottom"
  )

plot_grid(death_tiles, death_cols, align = "h", rel_widths = c(1, 0.2))
```

The minimum plot date (and the max) is the same as for cases, so it is possible to see the delay in deaths since the cases.

LAs are ordered differently as those of cases.  I am in no position to speculate why at this time!

I can see an interesting split in this death data... LAs who had high daily deaths in the first wave, had lower in the second; but those with fewer in the first suffered badly in the second.  Ordering by date of highest daily count does not present any correlation with total number of deaths.

