---
title: "Analysing Historical Weather"
description: "Understanding a little historical weather"
output: html_document
category: weather
tags: [r, weather]
layout: post
---

```{r echo = FALSE, include=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(ggthemes)
library(patchwork)
```

```{r knit_options, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
```

## Collecting Data

To start with, we can use the MetOffice's <a href="https://www.metoffice.gov.uk/research/climate/maps-and-data/historic-station-data" targe="_blank">Historic station data</a>.

There aren't too many stations offering data, so _Eastbourne_ is closest to me (known as the "Sunshine Coast", so I expect data to be a bit different to what I perceive north of the downs!).

The data is a text file, located here, <https://www.metoffice.gov.uk/pub/data/weather/uk/climate/stationdata/eastbournedata.txt>.

### Understanding the Data

First import the data. A quick visual shows that we have some introductory text - which we are unable to import (hence using `skip`). `fill` is required to cater for missing values.

It turned out that there are 2 rows of headers, and the 2nd row starting with blanks confused the import - therefore we read it for the headers then add those to the main data.

We also have to deal with the 'Provisional' text in a column that is not labelled.

```{r fetch, echo = TRUE, cache = TRUE}
.fetch <- function(url) {
  headers <- scan(
    file = url,
    skip = 5,
    nlines = 1,
    what = character()
  )

  data <- read.table(
    url(url),
    header = FALSE,
    fill = TRUE,
    skip = 7,
    col.names = c(headers, "isProvisional")
  )

  return(data)
}

rawEbn <- .fetch( "https://www.metoffice.gov.uk/pub/data/weather/uk/climate/stationdata/eastbournedata.txt") #nolint

str(rawEbn)
```

Cool. We have _`r nrow(rawEbn)`_ observations and _`r ncol(rawEbn)`_ variables.

### Getting Tidy Data

The docs detail some additional characters, that we don't need. The following characters are removed from the data:

- `#`, use of a different sensor
- `*`, 'estimated'
- `---`, more than 2 days missing data in the month

```{r tidy_dataset, echo = TRUE}
.tidyData <- function(rawData) {
  d <- rawData %>%
    select(
      year = yyyy,
      month = mm,
      maxTemp = tmax,
      minTemp = tmin,
      airFrost = af,
      rain = rain,
      sun = sun
    ) %>%
    mutate(
      across(
        !c(year, month),
        ~ stringr::str_replace_all(.x, c("\\*" = "", "#" = "", "---" = ""))
      ),
      across(!c(year, month), ~ as.double(.x)),
      across(c(year, month), ~ as.integer(.x))
    )

  # only get years with 12 months of data (to prevent totals being skew)
  minYear <- d %>%
    filter(month == 1) %>%
    slice_min(year)

  maxYear <- d %>%
    filter(month == 12) %>%
    slice_max(year)

  d <- d %>% filter(between(year, minYear$year, maxYear$year))

  return(d)
}

ebn <- .tidyData(rawEbn)
```

Now we have our tidy data, we can ask some questions...

### Asking questions

#### Is there a trend in annual rainfall?

```{r rainfall_total_plot}
.plotTotalRain <- function(data) {
  data %>%
    group_by(year) %>%
    summarise(
      totalRain = sum(rain)
    ) %>%
    ggplot(aes(year, totalRain)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(
      x = "",
      y = "Annual Rain /mm",
      caption = "Source: MetOffice Historic Station Data"
    ) +
    theme_economist_white()
}

.plotTotalRain(ebn)
```

Annual rainfall is pretty random, but there is a notable trend downwards.

#### Which month has maximum and minimum rainfall?

```{r rainfall_max_min}
maxRainfall <- ebn %>% slice_max(rain)
minRainfall <- ebn %>% slice_min(rain)
```

- Minimum rainfall month was in _`r paste0(month.name[minRainfall$month], " ", minRainfall$year)`_, and measured _`r minRainfall$rain`mm_.
- Maximum rainfall month was in _`r paste0(month.name[maxRainfall$month], " ", maxRainfall$year)`_, and measured _`r maxRainfall$rain`mm_.

#### How does monthly rainfall vary?

Whisker and box plots show median values (for each month) alongside upper and lower quartiles (50%, 25% and 75% respectively). 90% of data is covered between the whiskers. So at a median, 50% of values appear above the line and 50% below.

```{r monthly_rainfall_box}
.monthlyRain <- function(data) {
  gg <- data %>%
    mutate(monthAbb = month.abb[month]) %>%
    ggplot(aes(reorder(monthAbb, month), rain, group = monthAbb)) +
    geom_boxplot(fill = "deepskyblue3") +
    labs(
      x = "",
      y = "Monthly Rainfall /mm",
      caption = "Source: MetOffice Historic station data"
    ) +
    theme_economist_white()

  rainExtreme <- data %>%
    group_by(month) %>%
    summarise(
      minRain = min(rain),
      maxRain = max(rain),
      range = max(rain) - min(rain)) %>%
    slice_max(range)

  return(list(range = rainExtreme, plot = gg))
}

rainData <- .monthlyRain(ebn)

rainData$plot
```

Almost all months have had near zero rainfall. _`r month.name[rainData$range$month]`_ has the greatest range, from _`r rainData$range$minRain`mm_ to  _`r rainData$range$maxRain`mm_.

#### How does monthly sunshine vary?

```{r monthly_sun_box}
.plotMonthlySun <- function(data) {
  gg <- data %>%
    mutate(monthAbb = month.abb[month]) %>%
    ggplot(aes(reorder(monthAbb, month), sun, group = monthAbb)) +
    geom_boxplot(fill = "gold3") +
    labs(
      x = "",
      y = "Monthly Sunshine /hrs",
      caption = "Source: MetOffice Historic station data"
    ) +
    theme_economist_white()

  range <- data %>%
    group_by(month) %>%
    summarise(
      minSun = min(sun),
      maxSun = max(sun),
      range = max(sun) - min(sun)) %>%
    slice_max(range)
  
  minMax <- data %>%
    group_by(year, month) %>%
    summarise(
      min = min(sun),
      max = max(sun)
    )

  return(list(
    range = range, 
    min = data %>% slice_min(sun),
    max = data %>% slice_max(sun),
    plot = gg))
}

sunData <- .plotMonthlySun(ebn)

sunData$plot
```

- The least amount of sun in a WHOLE month was in _`r month.name[sunData$min$month]` `r sunData$min$year`_ with just _`r sunData$min$sun`hrs_.
- The most amount of sun was in _`r month.name[sunData$max$month]` `r sunData$max$year`_ with _`r sunData$max$sun`hrs_.
- _`r month.name[sunData$range$month]`_ has the greatest range of sun, which has had between _`r sunData$range$minSun`hrs_ and _`r sunData$range$maxSun`hrs_.

> These values include 'outlier' data (outside of the 90% likleyhood).  Although I am not presenting actual quartile data here, you can see from the boxplot above which month is most LIKELY to have a wide range of sun hours.
>
> For Eastbourne, this is _June_.

#### Are maximum / minimum temperatures 'related'?

```{r max_min_temp}
.maxMinTemp <- function(data) {
  data %>%
    mutate(monthAbb = month.abb[month]) %>%
    ggplot(aes(reorder(monthAbb, month), group = monthAbb)) +
    geom_boxplot(aes(y = minTemp), fill = "darkblue") +
    geom_boxplot(aes(y = maxTemp), fill = "darkred") +
    labs(
      x = "",
      y = "Temperature (\u00B0C)",
      caption = "Source: MetOffice Historic station data"
    ) +
    theme_economist_white()
}

(.maxMinTemp(ebn))
```

Yes.

#### Do years have varying temperature range?

The minimum minimum to the maximum maximum of each year.

```{r temp_range, fig.height=11, units="in"}
.tempRange <- function(data) {
  d <- data %>%
    group_by(year) %>%
    summarise(
      minT = min(minTemp),
      maxT = max(maxTemp),
      diff = maxT - minT
    )

  gg <- d %>%
    ggplot(aes(y = year, colour = diff)) +
    geom_segment(aes(x = minT, xend = maxT, yend = year), size = 3) +
    geom_vline(aes(xintercept = mean((maxT + minT) / 2), linetype = "Mean"), size = 1.2) +
    scale_x_continuous(
      breaks = as.integer(c(
        seq(floor(min(data$minT)), 0, by = 5),
        seq(0, max(data$maxT), by = 5)))
    ) +
    scale_y_reverse() +
    scale_colour_gradient(
      low = "darkblue",
      high = "darkred"
    ) +
    scale_linetype_manual(values = c("Mean" = "dotdash")) +
    labs(
      x = "Temperature (\u00B0C)",
      y = "",
      colour = "Temp\nDifference",
      caption = "Source: MetOffice Historic station data"
    ) +
    theme_economist_white() +
    theme(
      legend.position = "bottom"
    )

  return(list(plot = gg, data = d))
}

temps <- .tempRange(ebn)

(temps$plot)

tempsData <- temps$data

minTemps <- tempsData %>% slice_min(diff)
maxTemps <- tempsData %>% slice_max(diff)
```

- Minimum temperature range is _`r minTemps$diff`°C_ in _`r minTemps$year`_, with a minimum of _`r minTemps$minT`°C_ and maximum of _`r minTemps$maxT`°C_.
- Maximum temperature range is _`r maxTemps$diff`°C_ in _`r maxTemps$year`_, with a minimum of _`r maxTemps$minT`°C_ and a maximum of _`r maxTemps$maxT`°C_.

## Create a Play-Thing

### All Weather

Remember that temperatures are the mean daily maximum/minimum for the month. Rain, sun and frost are the sum of mm, hours, days (resp).

> UPDATE: It seems that I cannot render to HTML if I connect them all and make them interactive :o(  So here they all are static...

```{r all_weather, warning=FALSE, fig.height=12}
.allWeather <- function(data) {
  totals <- #crosstalk::SharedData$new(
    data %>%
      group_by(year) %>%
      summarise(
        minTempYear = min(minTemp),
        maxTempYear = max(maxTemp),
        totalRain = sum(rain),
        totalSun = sum(sun),
        totalDaysFrost = sum(airFrost)
#      ),
#    key = ~year
  )

  .totalsPlot <- function(column, colour) {
    totals %>%
      ggplot(aes(year, .data[[column]])) +
      geom_point(colour = colour) +
      geom_smooth(method = "loess") +
      scale_x_continuous(expand = expansion(c(0, 0))) +
      labs(
        x = "",
        caption = "Source: MetOffice Historic station data") +
      theme_economist_white()
  }

  totalRain <- .totalsPlot("totalRain", "deepskyblue3") +
    labs(y = "Total Rain (mm)")
  minTemps <- .totalsPlot("minTempYear", "darkblue") +
    labs(y = "Min Temp (\u00B0C)")
  maxTemps <- .totalsPlot("maxTempYear", "darkred") +
    labs(y = "Max Temp (\u00B0C)")
  totalSun <- .totalsPlot("totalSun", "gold3") +
    labs(y = "Total Hours Sun")
  totalDaysFrost <- .totalsPlot("totalDaysFrost", "steelblue2") +
    labs(y = "Total Days Frost")

#  crosstalk::bscols(
  gridExtra::grid.arrange(
    (minTemps),
    (maxTemps),
    (totalRain),
    (totalSun),
    (totalDaysFrost),
    nrow = 5)
#    widths = 12
#  )
}

.allWeather(ebn)
```
