---
title: "Chocolate Ratings"
description: "Tidy Tuesday from 2022-01-18, chocolate ratings."
output: html_document
category: tidytuesday
tags: [r, tidytuesday]
layout: post

# This file is also first use of `|>` pipe (over dplyr's `%>%`).  Expect mishaps.
---

`tidytuesday's` _Chocolate Ratings_ from 18th January 2022, <https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-01-18>.

Using the data provided, we can investigate a few attributes of chocolate, including its rating vs characteristics; source location and manufacturer.

```{r knit_options, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
```

```{r setup}
library(dplyr)
library(ggplot2)
library(ggthemes)

theme_set(theme_minimal())
#theme_update(
#  text = element_text(size = 10)
#)
```

First, we import the data and visualise the type of data that we have.

```{r get-data}
# https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-01-18/readme.md

choco_raw <- readr::read_csv(
  'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-01-18/chocolate.csv')

summary(choco_raw)
```

## Which characteristics are used to describe the chocolate?

```{r characteristics}
memorable_words <- choco_raw |>
  select(memorable = most_memorable_characteristics) |>
  tidytext::unnest_tokens(word, memorable) |>
  count(word, sort = TRUE)

total_memorable <- sum(memorable_words$n)
total_chocolates <- nrow(choco_raw)

memorable_words |>
  slice_head(n = 20) |>
  mutate(word = stringr::str_to_title(word)) |>
  mutate(word = reorder(word, n)) |>
  ggplot(aes(n, word)) +
  geom_col(fill = "chocolate4") +
  labs(
    title = "Top 20 memorable characteristics of the chocolate",
    subtitle = paste0("Total characteristics: ", total_memorable, " for ", total_chocolates, " chocolates"),
    x = "Number of Mentions", 
    y = NULL,
    caption = "#tidytuesday (2022-01-08) | Graph by @yukkybananas"
) +
  ggthemes::theme_hc()
```

## Are these characteristics related to the review given?

```{r char_review}
# wrangle the data for the plot
choc_reviews <- choco_raw |> 
  select(rating, memorable = most_memorable_characteristics) |>
  mutate(memorable = strsplit(memorable, ",")) |>
  tidyr::unnest(memorable) |>
  count(memorable, rating, sort = TRUE) |>
  slice_head(n = 30) |>
  mutate(memorable = factor(stringr::str_to_title(memorable))) |>
  mutate(memorable = reorder(memorable, n))

# to provide our labels
tally <- choc_reviews |>
  group_by(rating) |>
  summarise(n = sum(n)) |>
  arrange(rating)
  
choc_reviews |>
  ggplot(aes(x = rating, y = memorable)) +
  geom_hline(aes(yintercept = memorable), colour = "grey") +
  geom_point(aes(fill = n, size = n), colour = "white", stroke = 2, shape = 21) +
  colorspace::scale_fill_continuous_sequential(
    palette = "BrwnYl"
  ) +
  scale_x_continuous(
    breaks = seq(2.75, 4, by = 0.25),
    labels = paste0(
      c("2.75\n(n = ", "3.0\n(n = ", "3.25\n(n = ", "3.5\n(n = ", "3.75\n(n = ", "4.0\n(n = "),
      tally$n,
      rep(")", 6)
    )
  ) +
  scale_size_continuous(
#    range = c(1, 20), 
    guide = "none") + # could also use `guides`
  labs(
    title = "Top 30 ratings for memorable characteristics",
    subtitle = "Sweet chocolate is lower rated than chocolate with cocoa",
    x = "Rating (n = Number of mentions)",
    y = NULL,
    fill = "Number\nof times\nmentioned",
    caption = "#tidytuesday (2022-01-08) | Graph by @yukkybananas"
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(
      margin = margin(t = 10)
    ),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank()
  )

#ggsave("figure/R/tidytuesday/20220118.png", width = 7, height = 7, dpi = 320, bg = "white")

rtings <- choco_raw %>% count(rating, sort = TRUE)
```

This plot suggestst that `3.0` is the most common rating (`r rtings |> filter(rating == 3.0) |> dplyr::select(n)` ratings), but `3.5` is most rated overall (`r rtings |> filter(rating == 3.5) |> select(n)` ratings).  It is likely that a lot of other characteristics have few ratings each at `3.5`.  People seem a bit indifferent to their chocolate!

## Is the mention of `cocoa` preferred high % or low %?

```{r popular_cocoa}
choco_percent <-
  choco_raw |>
  select(cocoa_percent) |>
  count(cocoa_percent, name = "All Bars")

choco_percent_cocoa <- 
  choco_raw |>
  select(cocoa_percent, memorable = most_memorable_characteristics) |>
  tidytext::unnest_tokens(word, memorable) |>
  filter(word == "cocoa") |>
  count(cocoa_percent, name = "Mentions Cocoa")

coef <- max(choco_percent$`All Bars`)

joined <- choco_percent |>
  full_join(choco_percent_cocoa, by = c("cocoa_percent")) |>
  mutate(
    cocoa_percent = as.integer(stringr::str_remove(cocoa_percent, "%")),
    pc = `Mentions Cocoa` / `All Bars`) |>
  tidyr::pivot_longer(!c(cocoa_percent, pc),  names_to = "type", values_to = "value") |>
  mutate(type = factor(type)) |>
  arrange(cocoa_percent, type)

joined |>
  ggplot(aes(x = cocoa_percent, y = value)) +
  geom_line(aes(colour = type), alpha = 0.6) +
  geom_point(data = joined |> filter(type == "Mentions Cocoa"), aes(y = pc * coef, shape = "% mentioning\n'cocoa'"), colour = "orange3", stroke = 3, shape = "-") +
  scale_x_continuous(
    breaks = seq(10, 100, by = 10),
    labels = scales::percent_format(scale = 1)
  ) +
  scale_shape_manual(values = c("% mentioning\n'cocoa'" = "-")) +
  scale_y_continuous(
    trans = "log10",
    sec.axis = sec_axis(
#      name = "Percent bars mentioning 'cocoa'",
      trans = ~ . / coef,
      labels = scales::percent_format(scale = 100)
    )
  ) +
  labs(
    title = "Cocoa content per bar",
    subtitle = "Mentions of cocoa tracks the number of bars\nwith that percentage of cocoa",
    x = "Cocoa %",
    y = "Number of bars",
    caption = "#tidytuesday (2022-01-08) | Graph by @yukkybananas",
    shape = NULL,
    colour = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

This is a strange plot - I agree!  Area just wasn't working, and line seems to be missing a few points...  `¯\_(ツ)_/¯`

- The percentage of cocoa in a bar, where _cocoa_ is a memorable characteristic, seems to follow the total number of available with that percentage of cocoa.
- `70%` cocoa is both the most available, and the most memorable with _cocoa_.
- There is even a bar with 100% cocoa (seriously??), but _cocoa_ was only memorable in `4.8%` of them!
- It's interesting that in general, percentage tends to peak when divisible by 5 (eg, 70%, 75%, 80%).  Perhaps this is rounded on behalf of consumers; or because (eg) 73% suggests accuracy that cannot be achieved; or because round numbers are actually more tasty; or something else...
- The highest proportion of memorable _cocoa_ to the number of bars available is `50%` at `56%`, `71%`, `73%` cocoa percent.

So - cocoa seems to be preferred around 70% - 75%.  Sounds a lot.  Testers may have been chocolate "connoisseurs"!  More chart(s) would be needed to discover if these are well rated. 