---
title: "Netflix Shows"
author: "Jo Faircloth"
output: html_document
category: tidytuesday
tags: [r, tidytuesday]
layout: post
---

`tidytuesday"s` _Netflix Shows_ from 20th April 2021, <https://github.com/rfordatascience/tidytuesday/blob/eedfa9fe5c39d973502f00b2574ba14bff7ed82b/data/2021/2021-04-20>.

Data sourced from [Kagle](https://www.kaggle.com/shivamb/netflix-shows?select=netflix_titles.csv).

```{r knit_options, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  root.dir = "/")
```

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggthemes)
library(patchwork)
```

```{r data}
netflix <- readr::read_csv("2021-04-20/netflix_titles.csv")
```

```{r colours}
colours <- c(red = "#e50914", black = "#222222", grey = "#303030", text = "#ffffff")

covidDate <- as.Date("2020-03-15")
disneyDate <- as.Date("2019-11-12")
```

## Monthly title releases

The number of titles released onto Netflix each month.

The launch of Disney+ apears to have provided a peak for new films on Netflix.

Covid-19 possibly led to a lower release rate for the following couple of years, but it is not too dis-similar to the years up to the launch of Disney+.

```{r plot_monthly_titles}
n <- netflix |>
  mutate(
    added = lubridate::mdy(date_added),
    added = as.Date(added),
    added_month  = lubridate::floor_date(added, unit = "month")) |>
  group_by(added_month) |>
  summarise(n = n())

ggTiles <- n |>
  ggplot() +
  geom_tile(aes(x = added_month, y = 1, fill = n, width = 35)) +
  scale_fill_gradient(low = colours["grey"], high = colours["red"]) +
  scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y",
    expand = expansion(c(0, 0))
  ) +
  labs(
    caption = "Visualisation: Jo Faircloth\nSource: Kaggle | #tidytuesday"
  ) +
  guides(
    fill = guide_colorbar(
      title = "Netflix titles added per month",
      title.position = "bottom",
      title.hjust = 0.5,
      barwidth = 20,
      barheight = 0.5,
      frame.colour = colours["text"]
    )
  ) +
  theme_void() +
  theme(
    axis.text.x = element_text(colour = colours["text"], size = 12),
    text = element_text(
      colour = colours["text"]
    ),
    legend.position = "bottom",
    legend.box.margin = margin(8, 0, 4, 0, "mm"),
    plot.background = element_rect(fill = colours["black"]),
    plot.margin = margin(3, 3, 2, 3, "mm"),
    plot.caption = element_text(
      colour = colours["text"],
      hjust = 1,
      size = 8
    )
  )

# covid
ggTiles <- ggTiles +
  geom_vline(
    aes(xintercept = covidDate), 
    linetype = "dotted",
    size = 0.5, 
    colour = colours["text"],
    alpha = 0.8) +
  annotate(
    geom = "text",
    x = covidDate - 400,
    y = 1.2,
    label = "Covid-19:\nglobal lockdows\nbegan",
    colour = colours["text"]) +
  geom_curve(
    aes(
      x = covidDate - 365, 
      y = 1.12, 
      xend = covidDate,
      yend = 1.05),
    curvature = 0.3,
    colour = colours["text"],
    arrow = arrow(length = unit(2, "mm"))
  )

# disney
ggTiles <- ggTiles +
  geom_vline(
    aes(xintercept = disneyDate), 
    linetype = "dotted",
    size = 0.5, 
    colour = colours["text"],
    alpha = 0.8) +
  annotate(
    geom = "text",
    x = disneyDate - 400,
    y = 0.6,
    label = "Disney+ launched\nin the USA",
    colour = colours["text"]) +
  geom_curve(
    aes(x = disneyDate - 365, y = 0.65, xend = disneyDate, yend = 0.7),
    curvature = -0.3,
    colour = colours["text"],
    arrow = arrow(length = unit(2, "mm"))
  )

ggTiles
```


### <a href="https://raw.githubusercontent.com/misstickles/misstickles.github.io/main/figure/R/tidytuesday/20210420" target="_blank">Full size plot</a>


```{r eval=FALSE}
ggsave("figure/R/tidytuesday/20210420.png", plot = ggTiles, bg = colours["black"], dpi = 300, width = 8, height = 5)
```



