---
title: "Board Games"
author: "Jo Faircloth"
output: html_document
category: tidytuesday
tags: [r, tidytuesday]
layout: post
---

`tidytuesday"s` _Board Gamess_ from 25th January 2022, <https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-01-25>.

Data sourced from [Kagle](https://www.kaggle.com/datasets/jvanelteren/boardgamegeek-reviews?select=2022-01-08.csv).

```{r knit_options, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  root.dir = "/")
```

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggthemes)
#library(patchwork)
```

```{r data}
ratings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-01-25/ratings.csv')

details <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-01-25/details.csv')
```

```{r colours}
colours <- c("")
```

## Mean Rating Over the Years

```{r yearly_ratings}
library(dygraphs)
library(xts)

details |>
  left_join(ratings, by = "id")

yearly <- ratings |>
  select(year, average) |>
  group_by(year) |>
  summarise(avg = mean(average), n = n())

yearly |>
  filter(year <= 2022 & year >= 1920) |>
  ggplot(aes(x = year, y = avg)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(
    breaks = seq(1920, 2022, 10),
    expand = expansion(c(0.01, 0.01))) +
  labs(
    x = NULL
  ) +
  theme_minimal()
```

```{css}
.dygraph-label { color: #F00; }
```

```{r yearly_digraph}
dg <- yearly |>
  filter(year >= 1920 & year <= 2022) |>
  mutate(year = as.Date(ISOdate(year, 1, 1))) |>
  select(year, avg)

ts <- xts(x = dg$avg, order.by = dg$year)

dygraph(ts, main = "Mean Game Ratings", ylab = "Mean Rating / Year") |>
  dyAxis("x") |>
  dyAxis("y", valueRange = c(0, 10)) |>
  dySeries("V1", label = "Rating") |>
  dyOptions(drawPoints = TRUE, labelsUTC = TRUE, fillGraph = TRUE, fillAlpha = 0.2, drawGrid = FALSE, colors = "green", stepPlot = TRUE) |>
  dyRangeSelector(strokeColor = "steelblue") |>
  dyCrosshair(direction = "vertical") |>
  dyHighlight(highlightCircleSize = 3, highlightSeriesBackgroundAlpha = 0.2, hideOnMouseOut = FALSE) |>
  dyEvent("2020-03-26", "UK Lockdown #1", labelLoc = "bottom") |>
  dyEvent("1939-09-1", "WWII begins", labelLoc = "bottom") |>
  dyEvent("1945-09-2", "WWII ends", labelLoc = "bottom") |>
  dyEvent("1957-01-01", "1957", labelLoc = "bottom")
```

Overall, until approximately 2007, mean ratings were steady around 5.8 / 6.0.  After 2007, people started rating their games higher.  The 2020 COVID-19 pandemic did not appear to change this trend.  People seemed to be able to enjoy their board games in the middle of World War 2 too.

**1957** has an 'unusual' drop in mean board game ratings.  At 3.93, it is 1.91 lower than 1956, 1.7 lower than 1958 and is the lowest rating in over 100 years.  [Some interesting things happeded](https://www.thepeoplehistory.com/1957.html) in 1957 - notably the first 'Frisbee' became available and there was a Flu pandemic in Asia.


### <a href="https://raw.githubusercontent.com/misstickles/misstickles.github.io/main/figure/R/tidytuesday/20210420" target="_blank">Full size plot</a>

```{r eval=FALSE}
ggsave("figure/R/tidytuesday/20210420.png", plot = ggTiles, bg = colours["black"], dpi = 300, width = 8, height = 5)
```